---
title: "Getting started with GenomicDistributions"
author: "Nathan Sheffield"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{1. Getting started with GenomicDistributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
# These settings make the vignette prettier
knitr::opts_chunk$set(results="hold", collapse=FALSE, message=FALSE)
#refreshPackage("GenomicDistributions")
#devtools::build_vignettes("code/GenomicDistributions")
#devtools::test("code/GenomicDistributions")
```

If you have a set of genomic ranges, the GenomicDistributions R package can help you with some simple visualizations. Currently, it can produce two kinds of plots: First, the *chromosome distribution plot*, which visualizes how your regions are distributed over chromosomes; and second, the *feature distribution plot*, which visualizes how your regions are distributed relative to a feature of interest, like Transcription Start Sites (TSSs).

## Install

Install `GenomicDistributions` like this:

```{r, eval=FALSE}
devtools::install_github("databio/GenomicDistributions")
```

## Initialization

Start by loading up the package and getting your query set of regions as a GenomicRanges object. I've included an example bed file to demonstrate how these plots look. You can load it up like this:

```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
library("GenomicDistributions")
queryFile = system.file("extdata", "vistaEnhancers.bed.gz", package="GenomicDistributions")
query = rtracklayer::import(queryFile)
```

## Chromosome distribution plots

*Chromosome distribution plots* help you visualize how your regions are distributed across chromosomes. To produce these, you'll need to specify the chomosome lengths for your reference assembly. There are a few ways to do this.

For the common reference assemblies that I use (hg19, hg38, mm9, and mm10), I've included the metadata in the package. If you're working with one of these genomes, making a plot of the distribution across chromosomes takes just a couple of lines of code:

```{r chromosome-distribution}
# First, calculate the distribution:
x = aggregateOverGenomeBins(query, "hg19")

# Then, plot the result:
plotGenomeAggregate(x)
```

What if we want to do the same thing but on 2 query sets at the same time? No problem:

```{r Chromosome distribution plots with multiple region sets}
# Let's fudge a second region set by shifting the first one over 
query2 = GenomicRanges::shift(query, 1e6)
queryList = GRangesList(vistaEnhancers=query, shifted=query2)
x2 = aggregateOverGenomeBins(queryList, "hg19")
plotGenomeAggregate(x2)
```

These functions just do a naive binning across the genome. If you want to tweak the way the bins are handled, or use a different reference assembly, that's also possible and is only slightly more complicated. There are other functions you can use for that, which are outlined in another vignette.

## Feature distance distribution plots

*Feature distance distribution plots* will show you how your regions are distributed with respect to the nearest feature of interest. To illustrate, we'll use Transcription Start Sites (TSS) as our example feature of interest (but really, you can use any region set).

For TSS plots, since this is such a common use case, we can use a handy built-in function that does everything for us. It's just one line of code to check distances from query to your TSSs (for common genomes), and then a second line of code to plot those distances:

```{r tss-distribution, fig.cap="TSS plot. Distribution of query regions relative to TSSs", fig.small=TRUE}
# Calculate the distances:
TSSDist = TSSDistance(query, "hg19")

# Then plot the result:
plotFeatureDist(TSSDist, featureName="TSS")
```

This plot uses log-scale increasing bins to show how your regions are distributed. Now, let's make a similar plot with multiple region sets input:

```{r TSS plots with multiple region sets}

TSSDist2 = TSSDistance(queryList, "hg19")
plotFeatureDist(TSSDist2)

```

If you want to check distances to other features, that's no problem; `TSSDistance()` is really just a wrapper for the workhorse function, `featureDistanceDistribution()`. To show how this works, get some features you want to check the distance to. Here, let's just shift our query set by a normally distributed random number:

```{r Build features}
featureExample = GenomicRanges::shift(query, round(rnorm(length(query), 0,1000)))
```

Now, with these features, we just use the `featureDistanceDistribution` function to calculate the distances. This function using the fast rolling joins from `data.table` under the hood, so it complete very quickly. The result of this gets piped right into the plotting function as before:

```{r Calculate distances and plot, fig.cap="Feature plot. Distribution of query regions relative to arbitrary features", fig.small=TRUE}
fdd = featureDistanceDistribution(query, featureExample)
plotFeatureDist(fdd)
```
