---
title: "Full power GenomicDistributions"
author: "Nathan Sheffield"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{2. Full power GenomicDistributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
# These settings make the vignette prettier
knitr::opts_chunk$set(results="hold", collapse=FALSE, message=FALSE)
#refreshPackage("GenomicDistributions")
#devtools::build_vignettes("code/GenomicDistributions")
#devtools::test("code/GenomicDistributions")
```

# Full power GenomicDistributions

This vignette shows how to use GenomicDistributions of full-size data. It is pre-computed. 

```{r, echo=FALSE}
# These settings make the vignette prettier
knitr::opts_chunk$set(results="hold", collapse=FALSE, message=FALSE)
#refreshPackage("GenomicDistributions")
#devtools::build_vignettes("code/GenomicDistributions")
#devtools::test("code/GenomicDistributions")
```

# Introduction

Welcome to the GenomicDistributions benchmarking.

Here's what you need to have installed:

```{r, eval=FALSE}
BiocManager::install("ChIPpeakAnno")
# BiocManager::install("EnsDb.Hsapiens.v75") # hg19
BiocManager::install("EnsDb.Hsapiens.v86") # hg38
devtools::install_github("databio/GenomicDistributions")
```


```{r}
library(microbenchmark)
library(GenomicDistributions)
library(ChIPpeakAnno)
# library(EnsDb.Hsapiens.v75) # hg19
library(EnsDb.Hsapiens.v86) # hg38

## create annotation file from EnsDb or TxDb
annoData <- toGRanges(EnsDb.Hsapiens.v86, feature="gene")
annoData[1:2]
```

Let's retrieve a local path to an example bed file and use BiocFileCache to download it here if we don't already have it locally:

```{r}
setwd("long_vignettes")  # run this from GenomicDistributions/long_vignettes
path = getwd()
bfc = BiocFileCache::BiocFileCache(getwd())
rpath = function(url, bfc) {
	message("Downloading ", url)
	BiocFileCache::bfcrpath(bfc, url)
}
urls = c(
	"http://big.databio.org/example_data/bedbase_tutorial/bed_files/GSE91663_ENCFF316ASR_peaks_GRCh38.bed.gz", 
	"http://big.databio.org/example_data/bedbase_tutorial/bed_files/GSE105977_ENCFF617QGK_optimal_idr_thresholded_peaks_GRCh38.bed.gz",
	"http://big.databio.org/example_data/bedbase_tutorial/bed_files/GSE91663_ENCFF319TPR_conservative_idr_thresholded_peaks_GRCh38.bed.gz",
	"http://big.databio.org/example_data/bedbase_tutorial/bed_files/GSM2827350_ENCFF928JXU_peaks_GRCh38.bed.gz")
bedpaths = lapply(urls, rpath, bfc)
```

Read this file in and build a GenomicRanges object:

```{r}

df = lapply(bedpaths, data.table::fread)
gr1 = GenomicDistributions:::dtToGr(df[[1]], chr="V1", start="V2", end="V3")

grs = lapply(df, GenomicDistributions:::dtToGr, chr="V1", start="V2", end="V3")
grs = lapply(grs, sort)
queryList = GRangesList(grs)
```

Set up benchmarking functions:

```{r}
TSSdist = calcFeatureDist(queryList, sort(annoData))
p = plotFeatureDist(TSSdist, featureName="TSS", tile=TRUE)
print(p)

perc2 = calcPartitionsRef(queryList, "hg38", metric = "percent")
p = plotPartitionPercents(perc2)
print(p)


ep2 = calcExpectedPartitionsRef(queryList, "hg19")
plotExpectedPartitions(ep2)

gc2 = calcGCContentRef(queryList, "hg38")
plotGCContent(gc2)

```


