---
title: "Full power GenomicDistributions"
author: "Nathan Sheffield"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{2. Full power GenomicDistributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
# These settings make the vignette prettier
knitr::opts_chunk$set(results="hold", collapse=FALSE, message=FALSE)
#refreshPackage("GenomicDistributions")
#devtools::build_vignettes("code/GenomicDistributions")
#devtools::test("code/GenomicDistributions")
```

# Full power GenomicDistributions

This vignette shows how to use GenomicDistributions of full-size data. It is pre-computed. 

```{r, echo=FALSE}
# These settings make the vignette prettier
knitr::opts_chunk$set(results="hold", collapse=FALSE, message=FALSE)
#refreshPackage("GenomicDistributions")
#devtools::build_vignettes("code/GenomicDistributions")
#devtools::test("code/GenomicDistributions")
```

# Introduction

Welcome to the GenomicDistributions benchmarking.

Here's what you need to have installed:

```{r, eval=FALSE}
BiocManager::install("ChIPpeakAnno")
# BiocManager::install("EnsDb.Hsapiens.v75") # hg19
BiocManager::install("EnsDb.Hsapiens.v86") # hg38
devtools::install_github("databio/GenomicDistributions")
```


```{r}
library(microbenchmark)
library(GenomicDistributions)
library(ChIPpeakAnno)
# library(EnsDb.Hsapiens.v75) # hg19
library(EnsDb.Hsapiens.v86) # hg38

## create annotation file from EnsDb or TxDb
annoData <- toGRanges(EnsDb.Hsapiens.v86, feature="gene")
annoData[1:2]
```

Let's retrieve a local path to an example bed file and use BiocFileCache to download it here if we don't already have it locally:

```{r}
path <- getwd()
bfc <- BiocFileCache::BiocFileCache(getwd())
url = "http://big.databio.org/example_data/bedbase_tutorial/bed_files/GSM2423312_ENCFF155HVK_peaks_GRCh38.bed.gz"
bedpath = BiocFileCache::bfcrpath(BiocFileCache::BiocFileCache(getwd()), url)
```

Read this file in and build a GenomicRanges object:

```{r}
df = data.table::fread(bedpath)
gr1 = GenomicDistributions:::dtToGr(df, chr="V1", start="V2", end="V3")
gr1
```

Set up benchmarking functions to compare:

```{r}
TSSdist = calcFeatureDist(sort(gr1), sort(annoData))
p = plotFeatureDist(TSSdist, featureName="TSS")
print(p)
```
