---
title: "Full-size data from GenomicDistributionsData"
author: "Michal Stolarczyk"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{3. Full-size data from GenomicDistributionsData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
# These settings make the vignette prettier
knitr::opts_chunk$set(results="hold", collapse=FALSE, message=FALSE)
```

# Introduction

The point of this vignette is to demonstrate the seamless usage of out compation package: `GenomicDistributionsData` as a source of reference data sets that are not included in `GenomicDistributions`.

## Motivation

`GenomicDistributions` package comes with build-in reference data sets to perform calculations for human hg19 genome. 

To use `GenomicDistributions` with other reference genomes you need to install our companion package with more full-size data: `GenomicDistributionsData`. It's currently hosted on our server.

```{r, eval=FALSE}
install.packages("http://big.databio.org/GenomicDistributionsData/GenomicDistributionsData_0.0.1.tar.gz", repos=NULL)
```

## `GenomicDistributionsData` contents

This package provides the following data:

```{r}
library(GenomicDistributions)
library(GenomicDistributionsData)
datasetListIQR = utils::data(package="GenomicDistributionsData")
datasetList = datasetListIQR$results[,"Item"]
datasetList
```

# Usage

With the package loaded we have access to the required files for more genomes, namely:

* hg38
* hg19
* mm10
* mm9

Let's calculate some statistics and create plot for an hg38 BED file example. First, let's download one and read into R workspace:

```{r}
download.file(url="http://big.databio.org/GenomicDistributionsData/GSE91663_ENCFF319TPR_conservative_idr_thresholded_peaks_GRCh38.bed", destfile="example_hg38.bed")
query = rtracklayer::import("example_hg38.bed", extraCols=c(signalValue = "numeric", pValue = "numeric", qValue = "numeric", peak = "integer"))
```
```{r, echo=FALSE}
file.remove("example_hg38.bed")
```


```{r chrom-plots-single, fig.cap="Chromosome distribution plots."}
# First, calculate the distribution:
x = calcChromBinsRef(query, "hg38")

# Then, plot the result:
plotChromBins(x)
```

The functions work with "hg38" as the reference. `GenomicDistributions` used required reference data sets from `GenomicDistributionsData` behind the scenes.

Importantly, using `GenomicDistributionsData` package you can perform calculations that were not possible before, since one of the inputs to `calcOpenSignal` function is a large matrix. Now, you just need to load in into the workspace, like this:

```{r specificity-plot-single, fig.height = 6, fig.cap="Specificity of chromatin accessibility across cell types."}
data(openSignalMatrix_hg38)
op = calcOpenSignal(query, openSignalMatrix_hg38)
plotOpenSignal(op)
```
